---
title: "Map of field sites with aridity gradient"
author: "Kian Kelly"
output:  github_document
#output: pdf_document
---

```{r setup, include=FALSE}
#Setup RMarkdown doc; not included in knit
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(fig.width = 6, fig.height = 4)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# Load Libraries

```{r load, message=FALSE, warning=FALSE}
library(GGally)
library(tidyverse)
library(scales)
library(network)
library(vroom)
library(phyloseq)
library(vegan)
library(patchwork)
library(ggnewscale)
library(ggforce)
library(geosphere)
library(rnaturalearth)
library(sf)
library(rnaturalearthdata)
library(dplyr)
library(raster)
library(purrr)
library(USAboundaries)
library(USAboundariesData)
library(ggspatial)


```

### Adapted from: <https://www.painblogr.org/2020-12-15-climate-change.html>

# 1: Download the files from the TerraClimate website

#### The TerraClimate project is basically a dataset containing climate station data which was interpolated on a global scale. It was the highest resolution dataset of it's kind. 

```{r}

#--- Download the files from the TerraClimate website ---#
# Precipitation
if (!file.exists('./ppt.nc')) {
  download.file(url = 'http://thredds.northwestknowledge.net:8080/thredds/fileServer/TERRACLIMATE_ALL/data/TerraClimate_ppt_2023.nc',
                destfile = 'ppt.nc')
}

# Evapotranspiration
if (!file.exists('./pet.nc')) {
  download.file(url = 'http://thredds.northwestknowledge.net:8080/thredds/fileServer/TERRACLIMATE_ALL/data/TerraClimate_pet_2019.nc',
                destfile = 'pet.nc')
}
```
#### These are netCDF files which are basically like a CSV but with multiple dimensions (ie climate data for multiple months). Here you can see the raster data for each month of the year. Rasters are basically objects containing values associated with coordinates like x and y or lat and long
```{r}

#--- Import the downloaded files ---#
# Precipitation
ppt <- stack(x = 'ppt.nc')

# Evapotranspiration
pet <- stack(x = 'pet.nc')

#--- Inspect ---#
# Precipitation
plot(ppt)
```

```{r}
# Evapotranspiration
plot(pet)
```

# 2: Calculate average Precipitation and Evapotranspiration for the world, crop for CA
#### You can use raster maths functions to calculate means for multiple months. 
```{r}
#--- Raster maths ---#
# Precipitation
ppt_mean <- calc(ppt, # RasterStack object
                 fun = mean, # Function to apply across the layers
                 na.rm = TRUE)

# Evapotranspiration
pet_mean <- calc(pet,
                 fun = mean, 
                 na.rm = TRUE)

```

## Crop for California
#### Here I crop coordinates for California, but you can use any coordinates you like.
```{r}
#--- Set the extent ---#
# Cut off all values outside California
ext <- extent(c(xmin = -124, xmax = -114, 
                ymin = 32, ymax = 42))



#--- Crop ---#
# Precipitation
ppt_mean_cal <- crop(x = ppt_mean, 
                 y = ext)

# Evapotranspiration
pet_mean_cal <- crop(x = pet_mean, 
                 y = ext)

#--- Inspect ---#
# Precipitation
plot(main = 'Precipitation',
     ppt_mean_cal)

```
```{r}
# Evapotranspiration
plot(main = 'Evapotranspiration',
     pet_mean_cal)
```

# 3: Calculate aridity by overlaying the rasters and log normalize

#### TerraClimate does not automatically calculate aridity. Aridity is the ratio of Precipitation to evapotransperation

```{r}

#--- Calculate aridity index ---#
# Precipitation (ppt) / Evapotranspiration (pet)
aridity_index <- overlay(x = ppt_mean_cal, # Raster object 1
                         y = pet_mean_cal, # Raster object 2
                         fun = function(x, y){return(x / y)}) # Function to apply

plot(main = 'Aridity index',
     aridity_index)
```

#### Here we convert the raster to a dataframe and log normalize for later plotting steps. I am using relative aridity here since this is better for high resolution visualization. 
```{r}
#--- Convert raster to a matrix ---#
aridity_index_matrix <- rasterToPoints(aridity_index)

#--- Convert to the matrix to a dataframe ---#
aridity_index_df <- as.data.frame(aridity_index_matrix)


aridity_index_df$layer = log10(aridity_index_df$layer + 1)

```

# 4: Plot the first map

## Get California geometry and turn to sf object

#### You can use USAboundaries package to obtain the California geometry as I did here. If you are interested in another state you can modify this. the geometry is loaded in st format which is not compatible with ggplot so you can use st as sf to convert it. 

```{r}

# Get the geometry of California
ca <- us_states(states = "CA", resolution = "high")

ca_geom <- st_as_sf(ca)


```

## Read in shapefile for California Deserts

#### You can obtain these shapefiles for different climates using www.sciencebase.gov if you are intrested in adding park borders, etc.

```{r}
# Path to your shapefile
# source: https://www.sciencebase.gov/catalog/item/5835e1cae4b0d9329c801b7b
shapefile_path_moja <- "./ca_sections.shp"

# Read the shapefile
shapefile_mojave <- st_read(shapefile_path_moja)

# Now you can work with the shapefile object
# For example, you can view its attributes, plot it, etc.
head(shapefile_mojave)
```

#### Set a color pallette

```{r}
# Define the number of colors you want in the palette
num_colors <- 10000

# Generate the palette from white to red
palette <- colorRampPalette(c("beige", "forestgreen", "darkgreen"))(num_colors)



```

## Plot

#### the tricky part here is loading in the shapefiles. geom_sf lets you do this. geom_raster allows you to plot the aridity data. How this works is each object has associated coordinates which can be overlayed. 
```{r}

p <- ggplot() +
  geom_raster(data = aridity_index_df,
              aes(y = y, x = x, fill = layer)) +
  scale_fill_gradientn("Relative aridity", colours = palette) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 10),
        axis.title = element_blank(),
        panel.grid.major = element_line(linetype = 2, 
                                        size = 0.5,
                                        colour = 'lightblue'),
        panel.grid.minor = element_blank(),
        # Set background color to blue
        panel.background = element_rect(fill = "lightblue")) + 
  geom_sf(data = ca_geom, aes(), alpha = 0, colour = "black", linewidth = 1.25) +
  geom_sf(data = shapefile_mojave, aes(), alpha = 0, colour = "black", linewidth = .4) +
  coord_sf(ylim = c(32.5, 37),
           xlim = c(-121, -115)) + 
  guides(fill = guide_colourbar(reverse = TRUE, title = "Relative aridity",
                                title.position = "top",
                                title.theme = element_text(size = 12)),
         colour = guide_legend(reverse = TRUE, title = "Relative aridity", title.position = "top")) 

p
```

# 5: Plot map with compass, scale, and metadata

#### Here I just made a simple excel sheet with X and Y coordinates for Lat and Long for each of my sites to plot coordinates. I also use ggspatial to add a compas and scalebar. 
```{r}

meta <- read_tsv("./Site_metadata.tsv")

meta$Site <- factor(meta$Site, levels = c("TP", "AB", "CIMA", "GMT","ODLO"))

p +
  ggspatial::annotation_scale(
    location = "bl",
    bar_cols = c("black", "white"),
    text_family = "ArcherPro Book"
  ) +
  ggspatial::annotation_north_arrow(
    location = "tl", which_north = "true",
    pad_x = unit(0.1, "in"), pad_y = unit(0.1, "in"),
    style = ggspatial::north_arrow_orienteering(
      fill = c("black", "white"),
      line_col = "black",
    )
  ) + geom_text(x = -116.5, y = 34.8, aes(label = "Mojave"), vjust = -0.5)  + geom_text(x = -115.3, y = 32.75, aes(label = "Colorado"), vjust = -0.5) +
  geom_point( data=meta, aes(x=Lattitude, y=Longitude, color=Site)) +
  scale_color_manual(name="Geographic Area", values= c("#F535AA", "red3", "green","blue","orange", "pink")) 
```
#### Save
```{r}

ggsave(filename = './art_map.png', plot = last_plot(), device = 'png', width = 10, height = 6, dpi = 300)
```
